---
title: "RSM 384 Group Project - NBA Defense and Size"
author: AliReza Rezvan, Melody Ashgriz, Jacob Wright, Jeffrey Chu, Tony Wang
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

```{r, include = FALSE}
install.packages("tidyverse")
install.packages("RCurl")
install.packages("corrplot")
install.packages("dplyr")
```

```{r, include=FALSE}
library(tidyverse)
library(RCurl)
library(corrplot)
library(dplyr)
library(stats) #for LOESS
library(nnet) #for Multinomial logit
```

## Data Collection: Merging NBA Draft Combine Data and 1996-2024 Advanced Defence Statistics

**NBA Combine (2000-2024)**

```{r}
url <- getURL("https://raw.githubusercontent.com/jeffreychu48/RSM384/refs/heads/main/NBA_draft_combine.csv")
nba.combine <- read.csv(text = url)
nba.combine <- separate(nba.combine, col = PLAYER, into = c("Last Name", "First Name"), sep = ", ")
nba.combine$Player <- paste(nba.combine$`First Name`, nba.combine$`Last Name`, sep = " ")
nba.combine <- nba.combine %>%
  mutate(WGT = as.numeric(WGT))
nba.combine <- nba.combine %>% select (Player, HGT, WGT, BAR, STNDRCH, PAN, LPVERT, SPRINT, LANE)
colnames(nba.combine)[1] = "PLAYER"
```

**1996-2024 NBA Individual Player Advanced Defensive Stats**

```{r}
url <- getURL("https://raw.githubusercontent.com/jeffreychu48/RSM384/refs/heads/main/1996-2024%20NBA%20Player%20Advanced%20Defensive%20Stats.csv")
nba.9624.player.adv.def <- read.csv(text = url)
```

[Creating a Statistical Minimum (Total Minutes \> 1500) and removing Outliers]{.underline}

```{r}
nba.9624.player.adv.def <- nba.9624.player.adv.def %>%
  mutate(GP = as.numeric(GP), MIN = as.numeric(MIN), DEFRTG = as.numeric(DEFRTG)) %>%
  mutate(TOTALMIN = GP*MIN, TOTALMIN = as.numeric (TOTALMIN)) %>% 
  filter(TOTALMIN > 1500, YEAR < "2024", !PLAYER == "Aaron McKie", !PLAYER == "A.C. Green")
```

**Merging the Two Datasets**

```{r}
nba.9624.merged <- nba.9624.player.adv.def %>% merge(nba.combine, by = "PLAYER", all = FALSE)

#removing N/A variables from data
nba.9624.logit <- na.omit(nba.9624.merged)  
nba.9624.logit <- nba.9624.logit %>% select(PLAYER, DEFRTG, HGT, WGT, BAR, PAN, LPVERT, SPRINT)
```

## Part 1: How has the emphasis on [physical attribute] for NBA’s Top Defenders been changing over time?

### GLM Model

[Setup: Creating New Measures for Top and Bottom 10% of defenders]{.underline}

```{r}
#top and bottom 10% of defenders 
top_10_def <- quantile(nba.9624.logit$DEFRTG, 0.1,)
bottom_10_def <- quantile(nba.9624.logit$DEFRTG, 0.9,)

## creating new variables to the dataset 
nba.9624.logit <- nba.9624.logit %>%
          mutate(ABOVE_DEFNDR = ifelse(DEFRTG<mean(DEFRTG), 1, 0), ELITE = ifelse(DEFRTG<=top_10_def, 1, 0), BAD = ifelse(DEFRTG>=bottom_10_def, 1, 0))
```

**Logistic Regression (GLM)**

*Based on Defensive Rating, taking into account a player's: Height (HGT), Weight (WGT), Wingspan to Height Ratio (BAR), Hand Size - Length x Width (PAN), Max Vertical (LPVERT), 3/4 Basketball Court Sprint (SPRINT), and Agility around the Basketball Key (LANE)*

[Above Average Defenders]{.underline}

```{r}
#Logistic regression
fit.average <- glm(ABOVE_DEFNDR ~ HGT + WGT + BAR + PAN + LPVERT + SPRINT, data = nba.9624.logit, family = "binomial")    

summary(fit.average)
```

[Elite Defenders]{.underline}

```{r}
#Logistic regression
fit.elite <- glm(ELITE ~ HGT + WGT + BAR + PAN + LPVERT + SPRINT, data = nba.9624.logit, family = "binomial")    

summary(fit.elite)
```

[Bad Defenders]{.underline}

```{r}
#Logistic regression
fit.bad <- glm(BAD ~ HGT + WGT + BAR + PAN + LPVERT + SPRINT, data = nba.9624.logit, family = "binomial")    

summary(fit.bad)
```

### Scatterplots

**Scatterplot Setup**

```{r}
#Probability of being an Above Average Defender
nba.9624.logit <- nba.9624.logit %>% 
  mutate(expected_aboveavg = fitted(fit.average)) 
```

```{r}
#Probability of being an Elite Defender
nba.9624.logit <- nba.9624.logit %>% 
  mutate(expected_elite = fitted(fit.elite)) 
```

```{r}
#Probability of being an Bad Defender
nba.9624.logit <- nba.9624.logit %>% 
  mutate(expected_bad = fitted(fit.bad)) 
```

```{r}
# Reshape data into Unpivoted format
nba.9624.unpivoted <- nba.9624.logit %>%
  pivot_longer(cols = starts_with("expected_"), 
               names_to = "category", 
               values_to = "value")
```

**Scatterplots: Defence and Hand Size (PAN)**

```{r}
# Create scaterplot
ggplot(nba.9624.unpivoted, aes(x = PAN, y = value, color = category, shape = category)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(
    values = c("expected_aboveavg" = "blue", "expected_elite" = "green", "expected_bad" = "red"),
    labels = c("expected_aboveavg" = "Above Average", "expected_elite" = "Elite", "expected_bad" = "Bad*")) +
  scale_shape_manual(
    values = c("expected_aboveavg" = 16, "expected_elite" = 15, "expected_bad" = 17),
    labels = c("expected_aboveavg" = "Above Average", "expected_elite" = "Elite", "expected_bad" = "Bad*")) +
  labs(x = "PAN", y = "Probability", color = "Defender Category", shape = "Defender Category") +
  ggtitle("Probability of Defender Type based on Hand Size (PAN)")
```

**Scatterplots: Defence and Plus Wingspan (BAR)**

```{r}
# Create scaterplot
ggplot(nba.9624.unpivoted, aes(x = BAR, y = value, color = category, shape = category)) +
  geom_point() +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_manual(
    values = c("expected_aboveavg" = "blue", "expected_elite" = "green", "expected_bad" = "red"),
    labels = c("expected_aboveavg" = "Above Average", "expected_elite" = "Elite*", "expected_bad" = "Bad")) +
  scale_shape_manual(
    values = c("expected_aboveavg" = 16, "expected_elite" = 15, "expected_bad" = 17),
    labels = c("expected_aboveavg" = "Above Average", "expected_elite" = "Elite*", "expected_bad" = "Bad")) +
  labs(x = "BAR", y = "Probability", color = "Defender Category", shape = "Defender Category") +
  ggtitle("Probability of Defender Type based on Plus Wingspan (BAR)")
```

## Part 2: How does Physical Attributes correlate with Defensive Rating?

### Average (ALIREZA):

All Players, Not Top 10 % (2004 - 2023 except for PAN which is 2010 onwards).

Acquire non-top 10% players:
```{r}
nba.90percent <- nba.9624.merged %>%
  group_by(YEAR) %>%
  filter(DEFRTG > quantile(DEFRTG, 0.1, na.rm = TRUE)) %>%
  ungroup()
```

Calculate yearly averages for the remaining 90% of players:
```{r}
nba.yearly.averages <- nba.90percent %>%
  group_by(YEAR) %>%
  summarise(
    avg_HGT = mean(HGT, na.rm = TRUE),
    avg_BAR = mean(BAR, na.rm = TRUE),
    avg_PAN = if (all(is.na(PAN))) NA else mean(PAN, na.rm = TRUE),
    avg_LPVERT = mean(LPVERT, na.rm = TRUE)
  )

head(nba.yearly.averages)
```

Trend of Average Height for Bottom 90% of players according to DEF RTG (Year 2004 onwards):
```{r}
nba.height.trend <- nba.yearly.averages %>%
  filter(YEAR >= 2004)

library(ggplot2)

ggplot(nba.height.trend, aes(x = YEAR, y = avg_HGT)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(
    title = "Trend of Average Height of NBA Players (Non-Top 10% Defenders) from 2004 Onward",
    x = "Year",
    y = "Average Height (avg_HGT)"
  ) +
  theme_minimal()
```
Trend of Average BAR for Bottom 90% of players according to DEF RTG (Year 2004 onwards):
```{r}
nba.bar.trend <- nba.yearly.averages %>%
  filter(YEAR >= 2004)

ggplot(nba.bar.trend, aes(x = YEAR, y = avg_BAR)) +
  geom_line(color = "purple") +
  geom_point(color = "purple") +
  labs(
    title = "Trend of Average Wingspan-to-Height Ratio (BAR) of NBA Players (Non-Top 10% Defenders) from 2004 Onward",
    x = "Year",
    y = "Average BAR"
  ) +
  theme_minimal()
```

Trend of Average PAN for Bottom 90% of players according to DEF RTG (Year 2010 onwards):
```{r}
nba.pan.trend <- nba.yearly.averages %>%
  filter(YEAR >= 2010)

ggplot(nba.pan.trend, aes(x = YEAR, y = avg_PAN)) +
  geom_line(color = "green") +
  geom_point(color = "green") +
  labs(
    title = "Trend of Average Hand Size (PAN) of NBA Players (Non-Top 10% Defenders) from 2010 Onward",
    x = "Year",
    y = "Average PAN"
  ) +
  theme_minimal()
```

Trend of Average LPVERT for Bottom 90% of players according to DEF RTG (Year 2004 onwards):
```{r}
nba.lpvert.trend <- nba.yearly.averages %>%
  filter(YEAR >= 2004)

ggplot(nba.lpvert.trend, aes(x = YEAR, y = avg_LPVERT)) +
  geom_line(color = "orange") +
  geom_point(color = "orange") +
  labs(
    title = "Trend of Average Max Vertical (LPVERT) of NBA Players (Non-Top 10% Defenders) from 2004 Onward",
    x = "Year",
    y = "Average LPVERT"
  ) +
  theme_minimal()
```


### Top Defenders (JEFFREY):

**Creating a Time Series measure of the Physical Attributes of the Top 25 Defenders each year (2004-2023):**

```{r}
top.25.defrtg.per.year <- nba.9624.merged %>%
  mutate(YEAR = as.numeric(YEAR)) %>% 
  filter(YEAR >= 2004) %>% 
  group_by(YEAR) %>%
  arrange(desc(-DEFRTG)) %>%
  slice_head(n = 25) %>%
  ungroup() 

top.25.defrtg.per.year %>% 
  select(PLAYER, YEAR, DEFRTG, HGT, WGT, BAR, PAN, LPVERT) %>% 
  arrange(desc(YEAR)) %>% 
  head(5)
```

[20 Year "Height" Trend]{.underline}

```{r}
ggplot(top.25.defrtg.per.year, aes(x = YEAR, y = HGT, color = DEFRTG)) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") + 
  scale_y_continuous(limits = c(68, 90)) + 
  geom_point() + 
  ggtitle("Time Trend of Height for the NBA's Top 25 Defenders")
```

[20 Year "BAR" Trend]{.underline}

```{r}
ggplot(top.25.defrtg.per.year, aes(x = YEAR, y = BAR, color = DEFRTG)) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") + 
  scale_y_continuous(limits = c(0.95, 1.15)) +   
  geom_point() +
  ggtitle("Time Trend of Plus Wingspan for the NBA's Top 25 Defenders")
```

[10 Year "PAN" Trend]{.underline}

(Note: Hand Measurements only started in 2010, so we shrink our visualization)

```{r}
ggplot(top.25.defrtg.per.year, aes(x = YEAR, y = PAN, color = DEFRTG)) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") + 
  scale_y_continuous(limits = c(60, 120)) + 
  scale_x_continuous(limits = c(2010, 2024)) +
  geom_point() + 
  ggtitle("Time Trend of Hand Size for the NBA's Top 25 Defenders")
```

[20 Year "LPVERT" Trend]{.underline}

```{r}
ggplot(top.25.defrtg.per.year, aes(x = YEAR, y = LPVERT, color = DEFRTG)) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") + 
  scale_y_continuous(limits = c(25, 45)) + 
  geom_point() + 
  ggtitle("Time Trend of Max Vertical for the NBA's Top 25 Defenders")
```
